Litrepl: Literate Paper Processor Promoting Transparency More Than Reproducibility

### 2025-01-01 paper/paper.tex
Update paper

Litrepl is implemented in Python as a command-line text processor utility. Its
primary function is to take a text document as input, process it according to
specified command line arguments and environment settings, and then output the
resulting document through its standard output.

@@

%lresult
Litrepl's ability to maintain interpreter sessions in the background is crucial
for enabling a Read-Eval-Print Loop (REPL) environment. These resources, shown
in Figure \ref{fig:resource-allocation}, are stored as files within an auxiliary
directory.

@@

%lignore
\begin{figure*}[hbt!]
  \centering
  \begin{minipage}{\textwidth}
  \begin{verbatim}
  document       ::= (code | result | ignore | text)*
  code           ::= (code-normal | code-comment)
  result         ::= (result-normal | result-comment)
  code-normal    ::= "\begin{MARKER}" text "\end{MARKER}"
  code-comment   ::= "% MARKER" text "% noMARKER"
  result-normal  ::= "\begin{result}" text "\end{result}"
  result-comment ::= "% result" text "% noresult"
  ignore         ::= "% ignore" text "% noignore"
  text           ::= ...
  \end{verbatim}
  \end{minipage}
  \caption{An illustrative grammar template for LaTeX documents where marker
  serves as a parameter configured via command-line arguments for each supported
  interpreter class.}
  \label{fig:document_representation}
\end{figure*}
%lnoignore

@@

Template grammars similar to the illustrative example above are encoded for
Markdown and LaTeX formats. Before each run, Litrepl calls Lark\cite{Lark} to
compile a customized parser and uses it to access the sections.

@@

TODO

>>>>

Litrepl is implemented as a command-line text processor utility. Its primary
function is to take a text document as input, process it according to specified
command line arguments and environment settings, and then output the resulting
document through its standard output.

@@

%lresult
Litrepl's ability to maintain interpreter sessions in the background is crucial
for enabling a Read-Eval-Print Loop (REPL) environment. The associated
resources, shown in Figure \ref{fig:resource-allocation}, are stored as files
within an auxiliary directory.

@@

%% ignore %%
\begin{figure*}[hbt!]
  \centering
  \begin{minipage}{\textwidth}
  \begin{verbatim}
  document       ::= (code | result | ignore | text)*
  code           ::= (code-normal | code-comment)
  result         ::= (result-normal | result-comment)
  code-normal    ::= "\begin{MARKER}" text "\end{MARKER}"
  code-comment   ::= "% MARKER" text "% noMARKER"
  result-normal  ::= "\begin{result}" text "\end{result}"
  result-comment ::= "% result" text "% noresult"
  ignore         ::= "% ignore" text "% noignore"
  text           ::= ...
  \end{verbatim}
  \end{minipage}
  \caption{An illustrative grammar template for LaTeX documents where marker
  serves as a parameter configured via command-line arguments for each supported
  interpreter class.}
  \label{fig:document-representation}
\end{figure*}
%% noignore %%

@@

Template grammars similar to the illustrative example in Figure
\ref{fig:document-representation} are encoded for Markdown and LaTeX formats.
Before each run, Litrepl calls Lark\cite{Lark} to compile a customized parser
and uses it to access the sections.

Evaluation results are encoded back into the result sections, and the entire
document is printed. At this stage, certain conditions can be optionally
checked. First, setting \verb|--pending-exitcode| to a non-zero value instructs
Litrepl to report an error if a section takes longer than the timeout to
evaluate. Second, setting \verb|--exception-exitcode| directs Litrepl to detect
Python exceptions. Lastly, \verb|--irreproducible-exitcode| triggers an error if
the evaluation result doesn't match the text initially present in the result
section.

The last option implements the only formal check for aiding reproducibility that
Litrepl provides.

@@

\begin{comment}
% lai

Lets write the "Discussion" section. of the paper. Please check the grammar of
the following points and make a linked narrative out of them. Please use Latex
markup.

---

The technical decision to abstract interpreters using text streams has its
positive and negative sides. While simplicity is the obvious bonus, the negatvie
consequences can be groouped into (1) No parallel evaluation at the
communication level. Th interpreter gets locked until it finishes evaluation of
one snippet before it can proceed to another one. (2) Transferrable data type is
limited to text-only streams.

We argue that the lack of parallel execution at the communication level can be
compensated by using interpreter-specific parallelism where and if it is
supported. For example, Python programs can use all sort of subprocess
utilities while shell programs have full access to shell job control.

Restricting streams to text-only data type seems to be a more funcdamental
limitation. Litrepl lifts this limitation by supporting text-only document
formats. Both Latex and Markdown have means of including rich data but they
don't encode these data directly. Instead they rely on side channels: file
system or network for referring such objects. Litrepl shares both pros such as
good representation in well-known version control systems, and cons, e.g.
requirement to organize such a side-channel data transfer explicitly.

Another controversial technical solution - to transfer the whole document at
input and output - can have performance penalties. In our experience the system
works sufficienlty good for few-thousand line documents, but for larger
documents delays might become uncomfortable even on modern computers. Still we
prefer to keep this interface to make editor integration as simple as possible:
typical plugin can just pipe the whole document through the tool using few lines
of code.

In addition, we allow a more performant integrations which take responsibility
for pre-parsing and piping only the relevant parts of the document. For these
approach, Litrepl offers \verb|print-regexp| command printing the anchor regexp
in a few common formats.
% lnoai
\end{comment}

% lresult
The technical decision to abstract interpreters using text streams comes with
both advantages and disadvantages. A key advantage is simplicity. However, there
are notable negative aspects. Firstly, there is no parallel evaluation at the
communication level, meaning the interpreter is locked until it completes the
evaluation of one snippet before proceeding to the next. Secondly, the
transferable data type is restricted to text-only streams.

We argue that the lack of parallel execution at the communication level can be
mitigated using interpreter-specific parallelism, where supported. For instance,
Python programs can utilize various subprocess utilities, while shell programs
have full access to shell job control.

The restriction to text-only data types presents a more fundamental limitation.
Litrepl lifts this restriction by supporting text-only document formats. Both
LaTeX and Markdown incorporate rich data without encoding it directly, instead
relying on side channels, such as the file system or network, to refer to such
objects. Consequently, Litrepl shares the advantages of good representation in
well-known version control systems, and the disadvantages, such as the need to
explicitly organize side-channel data transfer.

Another controversial technical decision is transferring the entire document as
input and output, which can negatively impact performance. Our experience shows
that the system performs adequately for documents of a few thousand lines.
However, larger documents may experience uncomfortable delays, even on modern
computers. Despite this, we choose to maintain this interface because it
simplifies editor integration. A typical plugin can pipe the whole document
through the tool using just a few lines of code.

Moreover, we support more performance-oriented integrations that handle
pre-parsing and piping only relevant document parts. For these approaches,
Litrepl offers the \verb|print-regexp| command, which outputs the anchor regexp
in several common formats.
% lnoresult
